Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-06-28 hazelnut-popl17.tex

\emph{Structure editors} allow programmers to edit the tree structure of a program directly -- there are no syntax errors. This can have cognitive benefits, particularly for novice and end-user programmers (as evidenced by the popularity of structure editors like Scratch.) It also simplifies matters for tool designers, because they need not contend with malformed program text. %Structure editors for professional programmers are also under active development (e.g. \texttt{mbeddr}.)

This paper defines Hazelnut, a {structure editor} based on a small bidirectionally typed lambda calculus that takes this approach two steps further: Hazelnut's edit actions operate over terms that are not just syntactically well-formed but also statically and dynamically meaningful.
Na\"ively, this injunction against type errors would force programmers to construct terms in an unnatural ``outside-in'' manner. To avoid this problem, Hazelnut's {action semantics} automatically places terms that have a type that is incompatible with the expected type into a \emph{hole}, which safely defers the type compatibility check until the hole is \emph{finished}. Interestingly, this mechanism is related to a mechanism that arises in gradual type systems.

We intend Hazelnut as a minimal foundational calculus, rather than as an end-user tool itself. To that end, we show how Hazelnut's rich metatheory, which we have mechanized in Agda, guides the definition of an extension to Hazelnut. We also briefly discuss how Hazelnut's semantics lends itself to implementation as a functional reactive web application. Our reference implementation is written using \lstinline{js_of_ocaml}.

@@

\section{Hazelnut, Formally}
\label{sec:hazel}
\todo{revise text}\todo{add let and ifz}\todo{make eqn numbering work better}Hazelnut is based on the simply-typed lambda calculus extended with a single base type, $\tnum$. Its major constituents, introduced by example in the previous section, are:
\begin{itemize}
\item \textbf{H-types} and \textbf{H-expressions} (Sec. \ref{sec:holes}), which are terms with \emph{holes}. Holes mark subterms that are ``under construction.'' H-types classify H-expressions according to a {bidirectionally typed} static semantics.
\item \textbf{Z-types} and \textbf{Z-expressions} (Sec. \ref{sec:cursors}), which superimpose a single \emph{focus}\todo{cursor?} onto H-types and H-expressions (using Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.)
\item \textbf{Actions} (Sec. \ref{sec:actions}), which move the focus or modify the subterm in focus.

@@

\begin{equation}\label{r:moveparent-hole}
  \inferrule{ }{
    \performMove{
      \hhole{\zwsel{\hexp}}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hhole{\hexp}}
    }
  }
\end{equation}
\end{subequations}

\paragraph{Reachability}

We give the following theorem to demonstrate that the rules for movement do
indeed capture the process of moving the focus to any editable position
within a term. Intuitively, given two zippered expressions that are the
same up to the erasure of focus, its proof gives a list of composable
actions that, when applied to the first term produces the second.

As a consequence of applying Theorem \ref{thrm:actsafe} inductively at
every action in such a list, we couldn't hope to construct an ill-typed
term. Therefore the statement here matches the mutually recursive structure
of the bidirectional typing judgements.

\begin{theorem}[Reachability]\todo{why $\htau$ and not $\ztau$?}
\label{thrm:reach}
Both of the following hold:
\begin{enumerate}
  \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
   $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau}$ and $\removeSel{\zexp}
   = \removeSel{\zexp'}$ then $\exists L \in \mathit{list}
   ~\mathtt{action}$ such that $\mathit{iterate}~ L~ \zexp
   = \zexp'$.

  \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
   $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$ and $\removeSel{\zexp}
   = \removeSel{\zexp'}$ then $\exists L \in \mathit{list}
   ~\mathtt{action}$ such that $\mathit{iterate}~ L~ \zexp
   = \zexp'$.
\end{enumerate}
\end{theorem}

\subsubsection{Deletion} The $\aDel$ action replaces the selected subterm with an empty hole.

@@

Deletion within a Z-expression is similarly straightforward:
\begin{subequations}
\begin{equation}
  \inferrule{ }{
    \performSyn{\hGamma}{\zwsel{\hexp}}{\htau}{\aDel}{\zwsel{\hehole}}{\tehole}
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performAna{\hGamma}{\zwsel{\hexp}}{\htau}{\aDel}{\zwsel{\hehole}}
  }
\end{equation}
%\end{subequations}
% The base case turns into a hole:
%\begin{subequations}
% \begin{equation}
% \inferrule{ }{
%   \performDel{\zwsel{\hexp}}{\hehole}
% }
% \end{equation}
% The rules for the recursive ascription case is shown below. The other recursive cases are analagous:
% \begin{equation}
%   \inferrule{
%     \performDel{\zexp}{\zexp'}
%   }{
%     \performDel{\zexp : \htau}{\zexp' : \htau}
%   }
% \end{equation}
% \begin{equation}
%   \inferrule{
%     \performTyp{\ztau}{\aDel}{\ztau'}
%   }{
%     \performDel{\hexp : \ztau}{\hexp : \ztau'}
%   }
% \end{equation}
\end{subequations}

\paragraph{Deletability}\todo{do we care?}
The meaning of the deletion action is to remove the entire term in focus,
and its definition does not depend on the structure of the term being
deleted. Therefore, there is no theorem to state that acts as a checksum on
the rules analogous to Theorem \ref{thrm:reach} or \ref{thrm:construct},
since we could not have forgotten any rules corresponding to the structure
of the terms.

\subsubsection{Construction} The construction actions, $\aConstruct{\varphi}$, are used to construct terms of a shape indicated by $\varphi$ into the program at or around the focus.

@@

\paragraph{Finishability}\todo{do we care?}

As with deletion, since the meaning of the finish action does not depend on
the structure of the term being finished other than requiring that it must
be one of the two hole forms, there is no checksum theorem to state for
this action.

@@

\paragraph{Constructibility}\todo{why is this with the zipper rules not the
    construct rules?}

We give the following theorem to demonstrate that the rules for the action
$\aConstruct{\varphi}$ do indeed capture the process of building any term
of the form $\varphi$. Intuitively, its proof gives a list of composable
actions that, when applied one sequentially to an empty hole in focus,
build any well-typed expression.

As with Theorem \ref{thrm:reach}, the statement of this theorem must match
the mutually recursive structure of the bidirectional typing judgements.

\begin{theorem}[Constructability]\todo{why $\htau$ and not $\ztau$?}
\label{thrm:construct}
Both of the following hold:
\begin{enumerate}
  \item If $\hsyn{\hGamma}{\hexp}{\htau}$ then $\exists \zexp'$ $\exists
      L \in \mathit{list} ~\mathtt{action}$ such that
      $\mathit{iterate}~ L~ \zwsel{\hhole{}} = \zexp'$ and
      $\removeSel{\zexp'} = \hexp$

  \item If $\hana{\hGamma}{\hexp}{\htau}$ then $\exists \zexp'$ $\exists
      L \in \mathit{list} ~\mathtt{action}$ such that $\mathit{iterate}~
      L~ \zwsel{\hhole{}} = \zexp'$ and $\removeSel{\zexp'} = \hexp$
\end{enumerate}
\end{theorem}

Because actions are defined on the zippered form of expressions, we argue
that we can construct a term that is the same as the target up to the
erasure of focus rather than on the nose. Since any form inside a zippered
term may be in focus, there are linearly many such terms for a given
$\hexp$---Theorem \ref{thrm:construct} can be composed with
Theorem \ref{thrm:reach} to reach any one that might be desired in
particular.

\end{subequations}

>>>>

\emph{Structure editors} allow programmers to edit the tree structure of a program directly. This can have cognitive benefits, particularly for novice and end-user programmers (as evidenced by the popularity of structure editors like Scratch.) It also simplifies matters for tool designers, because they need not contend with malformed program text. %Structure editors for professional programmers are also under active development (e.g. \texttt{mbeddr}.)

This paper defines Hazelnut, a {structure editor} based on a small bidirectionally typed lambda calculus that takes two more steps in this direction: Hazelnut's edit actions operate over terms that are not only syntactically well-formed but also statically and dynamically meaningful.  
Na\"ively, enforcing an injunction on ill-typed edit states would force programmers to construct terms in an unnatural ``outside-in'' manner. To avoid this problem, Hazelnut's {action semantics} automatically places terms assigned a type that is incompatible with the expected type into a \emph{hole}, which safely defers the type compatibility check until the hole is \emph{finished}. Interestingly, this mechanism is related to a mechanism that arises in the semantics of gradual type systems.

We intend Hazelnut to serve as a principled foundational calculus, rather than as an end-user tool itself. To that end, we demonstrate how Hazelnut's rich metatheory, which we have mechanized in Agda, guides an extension to Hazelnut. We also  discuss how Hazelnut's semantics lends itself to implementation as a functional reactive program. Our reference implementation is written using \lstinline{js_of_ocaml}.

@@

\section{Hazelnut, Formally}
\label{sec:hazel}
\todo{revise text}\todo{add let and ifz}\todo{make eqn numbering work better}Hazelnut is based on the simply-typed lambda calculus extended with a single base type, $\tnum$. Its major constituents, introduced by example in the previous section, are:
\begin{itemize}
\item \textbf{H-types} and \textbf{H-expressions} (Sec. \ref{sec:holes}), which are terms with \emph{holes}. Holes mark subterms that are ``under construction.'' H-types classify H-expressions according to a {bidirectionally typed} static semantics.
\item \textbf{Z-types} and \textbf{Z-expressions} (Sec. \ref{sec:cursors}), which superimpose a single \emph{focus} onto H-types and H-expressions (using Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.)
\item \textbf{Actions} (Sec. \ref{sec:actions}), which move the focus or modify the subterm in focus.

@@

\begin{equation}\label{r:moveparent-hole}
  \inferrule{ }{
    \performMove{
      \hhole{\zwsel{\hexp}}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hhole{\hexp}}
    }
  }
\end{equation}

Reachability\todo{state reachability theorem}

\end{subequations}
\subsubsection{Deletion} The $\aDel$ action replaces the selected subterm with an empty hole.

@@

Deletion within a Z-expression is similarly straightforward:
\begin{subequations}
\begin{equation}
  \inferrule{ }{
    \performSyn{\hGamma}{\zwsel{\hexp}}{\htau}{\aDel}{\zwsel{\hehole}}{\tehole}
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performAna{\hGamma}{\zwsel{\hexp}}{\htau}{\aDel}{\zwsel{\hehole}}
  }
\end{equation}
%\end{subequations}
% The base case turns into a hole:
%\begin{subequations}
% \begin{equation}
% \inferrule{ }{
%   \performDel{\zwsel{\hexp}}{\hehole}
% }
% \end{equation}
% The rules for the recursive ascription case is shown below. The other recursive cases are analagous:
% \begin{equation}
%   \inferrule{
%     \performDel{\zexp}{\zexp'}
%   }{
%     \performDel{\zexp : \htau}{\zexp' : \htau}
%   }
% \end{equation}
% \begin{equation}
%   \inferrule{
%     \performTyp{\ztau}{\aDel}{\ztau'}
%   }{
%     \performDel{\hexp : \ztau}{\hexp : \ztau'}
%   }
% \end{equation}

\end{subequations}
\subsubsection{Construction} The construction actions, $\aConstruct{\varphi}$, are used to construct terms of a shape indicated by $\varphi$ into the program at or around the focus.

@@

@@

Constructibility\todo{state constructibility theorem}
\end{subequations}

### 2016-06-28 intro.tex

%Although programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program.

Every day, programmers use text-based tools to construct and manipulate programs.  
These programs are written in languages which are defined by a textual syntax. 
While textual syntax has proved to have practical utility, it also introduces some fundamental complexity into the programming process.

%Doesn't really start out with a bang, IMHO. I like to start papers out with a "big" problem.
%Most programming languages define a textual syntax. This allows programmers to use text editors and other standard text-based tools to construct and manipulate programs. While this is of substantial practical utility, it also introduces some fundamental complexity into the programming process.

Cognitively, it requires that programmers understand the subtleties of text parsing (e.g. the relative precedence of the various forms.) This can be particularly challenging for novice and end-user programmers.
One study~\cite{Altadmri:2015:MCI:2676723.2677258} of novice programmer error found that the most common of all errors was an syntactic error, specifically unbalanced parenthesis.
% Too much detail?

It also complicates matters for tool designers because they might be confronted with text that does not correspond to a well-formed, meaningful program.
By analyzing the same data set used in~\cite{6883030}, which contains 1,460 hours of fine-grained code editing logs collected from 21 programmers, we found that 44.2\% of the code edits resulted in syntactically malformed state of the source code.
%\todo{mention YoungSeok's data here}. 
These malformed states could be errors, but they also may be because the programmer is in the midst of a sequence of edit actions that leaves the text temporarily malformed or ill-typed.
%, or because the programmer has made a mistake. 
The language definition is silent about these situations, so it is difficult for tools to help programmers determine and execute a corrective course of action (e.g. by providing syntax highlighting~\cite{sarkar2015impact} and semantics-aware code completion services~\cite{Mooty:2010:CCC:1915084.1916348}.)

Some editors have developed \emph{ad hoc} workarounds for this problem, e.g. they might attempt to insert closing delimiters automatically, use whitespace to guess where a construct is likely to end, or recover from a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.\todo{citations?}) These heuristic methods are imperfect, complex and sometimes misleading.

>>>>

%Although programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program. 
% Programming languages, and therefore programs themselves, are rich mathematical structures. 
Well-typed programs are rich semantic structures, but human programmers typically construct and manipulate them only indirectly: by editing text that is parsed according to a textual syntax and then typechecked according to a static semantics. This indirection is of substantial practical utility, to be sure -- text editors and other text-based tools benefit from decades of development effort. However, it also introduces some fundamental complexity into the programming process.

% Every day, programmers use text-based tools to construct and manipulate programs.  
% These programs are written in languages which defined by a textual syntax. 
% While textual syntax has proved to have practical utility, it also introduces some fundamental complexity into the programming process.

% It also complicates matters for tool designers because tools are often confronted with text that does not correspond to a well-formed, meaningful program\todo{mention YoungSeok's data here}. This may be because the programmer is in the midst of a sequence of 
% edit actions that leaves the text temporarily malformed or ill-typed, or because the programmer has made a mistake. The language definition is silent about these situations, so it is difficult for tools to help programmers determine and execute a corrective course of action (e.g. by providing \todo{cite study on benefits of syntax highlighting}syntax highlighting and semantics-aware code completion services\todo{cite something?}.)
%Doesn't really start out with a bang, IMHO. I like to start papers out with a "big" problem.
%Most programming languages define a textual syntax. This allows programmers to use text editors and other standard text-based tools to construct and manipulate programs. While this is of substantial practical utility, it also introduces some fundamental complexity into the programming process.

First, it requires that programmers understand the various subtleties of textual syntax (e.g. the relative precedence of the various forms.) This can be particularly challenging for novice and end-user programmers \cite{Altadmri:2015:MCI:2676723.2677258}. %For example, one study of novice programmers found syntax errors (e.g. unbalanced parentheses) to be the most common class of errors.
% Too much detail?

Moreover, the fact that not all sequences of characters correspond to a well-formed, meaningful program complicates matters for tool designers. Tools are confronted with such text quite frequently -- by analyzing a dataset gathered by Yoon and Myers \cite{6883030} (which contains 1,460 hours of fine-grained edit logs collected from 21 programmers) we found that 44.2\% of code edits resulted in malformed edit states. Some additional percentage of edit states were well-formed but ill-typed (the dataset did not include enough information to determine the exact percentage.)  
Some of these edit states were due to programmer error, while others arose simply because the programmer was in the midst of a sequence of edit actions that left the text temporarily malformed or ill-typed. The language definition is silent about these edit states, so it is difficult to design useful editor services (e.g. syntax highlighting~\cite{sarkar2015impact} and semantics-aware code completion~\cite{Mooty:2010:CCC:1915084.1916348}) and report meaningful error messages\todo{citations}.
%, or because the programmer has made a mistake. 
% The syntax definition is silent about these situations, so it is difficult for tools to help programmers determine and execute a corrective course of action

Some editors have developed \emph{ad hoc} workarounds for this problem, e.g. they might attempt to insert closing delimiters automatically, use whitespace to guess where a construct is likely to end, use regular expressions to highlight malformed program text, or recover from a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.\todo{citations?}) These heuristic methods are imperfect, complex and sometimes misleading. % This may also help explain why the error messages emitted by parsers and typecheckers are often quite baroque\todo{citations}.
